\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-08-25 writing/manuscript.Rmd
wrote conclusion and added hypothesis figures

---
title: "How do resource abundance and stochasticity affect animals' spatial needs?"
subtitle: "A review with quantitative simulations and empirical modeling"
author: "Mezzini S. and Noonan M. J."
date: "`r Sys.Date()`"
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # double-space text
    - \usepackage{indentfirst} # indent first line of each paragraph
    - \usepackage[small]{titlesec} # smaller font for headings
    - \usepackage{wrapfig} # to wrap text around figures
    - \usepackage{caption}
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-20pt}

    # - \usepackage{gensymb}
    # - \usepackage{float}
    # - \usepackage{afterpage}
    # - \usepackage{ragged2e}
subparagraph: yes # needed for \usepackage[small]{titlesec}
# classoption:
# - twocolumn
output:
  bookdown::pdf_document2:
    number_sections: true
    toc: false
---

@@

# chuck options used to wrap figures:
## fig.align = "right":          align figures right
## out.width = "0.5\\textwidth": output width = 0.5 * the width of the text
## wrapfigure = list("R", 0.5)   wrap caption to right, caption width = 0.5 text width
```

<!-- place figures before the paragraph they are referred to so wrapfigure places them beside the paragraph -->

<!-- add section explaining what the paper details and what the objectives are? -->

@ciuti and @berger-tal: human presence in nature for other reasons than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979

Add @mueller_how_2011 for a study that accounted for variance in NDVI for animal movement

The habitat productivity hypothesis proposed by @harestad_home_1979 is supported, but it ignores many other important causes of movement [@seigle-ferrand_systematic_2021].

The ability to move allows animals to respond rapidly and continuously to heterogeneous and changing environments and conditions [@southwood_habitat_1977]. Animals may move for a variety of reasons, including searching for resources (e.g., food, nutrients, water, heat, a new den or nesting spot), a mate, or a new group. Movement also allows animals to escape predation or dangerous competition as well as defend resources and territories [@grant_whether_1993]. @nathan_movement_2008 conceptualized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

@@

such that the animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters, doing so with sufficient frequency and accuracy would not be simple. Rather than measuring changes in the various resources and needs directly, we can quantify animals' spatial needs as a response to average resource abundance and the variability around it, instead. This paper will illustrate how animals' spatial needs depend on average resource abundance and the variance around it. We present this idea through a quantitative theoretical framework and empirical examples where resource abundance is estimated using the Normalized Vegetation Index [NDVI; see @pettorelli_normalized_2011].

@@

@@

The favorableness of a patch or habitat often depends on a variety of factors, including resource abundance, competitive pressure, and predation risk. Since many of these factors' affect resource abundance unpredictably, it is useful to visualize it as a random variable^[In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$, $H$). Known values, such as realizations (i.e., known observations) of random variables, are indicated with lower-case letters (e.g., $r$, $h$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.]. Let $R$ indicate the amount of *resources* in a particular patch or area. For simplicity, we can assume $R$ can take any value between zero and infinity, and we can let $R$ follow a Gamma distribution with mean $\mu$ and variance $\sigma^2$ (which we can write as $R \sim \Gamma(\mu, \sigma^2)$)^[Although Gamma distributions are more often parameterized using shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ for simplicity. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean since as the mean approaches zero the variance also does: $\mu = 0 \iff k \theta = 0 \iff k \theta^2 = 0 \iff \sigma^2 = 0$.].

@@

The top row of figure \@ref(fig:5-5-sims) (constant $\mathbb V(R)$) illustrates the simulated spatial requirements of an animal living in habitats where the average resource abundance remains constant or changes over time. Note how the home range size, $H$, decreases nonlinearly as $\mathbb E(R)$ increases linearly (linear $\mathbb E(R)$, constant $\mathbb V(R)$), and how the 95% home range is more sensitive to changes in $\mathbb E(R)$ than the core home range. In regions where the average resource abundance changes over time (cyclical $\mathbb E(R)$, constant $\mathbb V(R)$ ), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mathbb E(R)$. However, estimates of spatial requirements based on estimated changes in $\mathbb E(R)$ should be interpreted carefully, since model error and unpredictable changes in $\mathbb E(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the expected spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mathbb E(R)$, constant $\mathbb V(R)$), since accurate long-term estimates of $\mathbb E(R)$ may be hard to produce, if not impossible. In cases where $\mathbb E(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often. Finally, in ecosystems where $\mathbb E(R)$ changes erratically and suddenly (erratic $\mathbb E(R)$, constant $\mathbb V(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

@@

Although the effect of resource abundance on animals' spatial needs is often recognized and accounted for, the variance in resource abundance is generally ignored or assumed to remain constant over time. However, resource abundance often varies due to multiple causes. Some causes are repetitive, predictable, or even well-known patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches), while others are infrequent, unpredictable, or poorly-understood events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018]. In this paper, we separate $\mathbb V(R)$ into heterogeneity and stochasticity. We refer to predictable or expected spatiotemporal changes as environmental **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable. In contrast, the quantity and quality of berries produced during a given year may be stochastic. Thus, $\mathbb V(R)$ is affected by both sources of variation, and changes in $\mathbb V(R)$ may predictable or not.

The effects of $\mathbb V(R)$ on animals' spatial requirements likely depend on the animals' needs. @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous) required exponentially less space than foxes that lived in more homogeneous regions. @nilsen_can_2005 found that the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality, while the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. ***ADD MORE EMPIRICAL EXAMPLES?*** However, the effects of environmental variance on animals' spatial needs remains largely undestudied and unaccounted for.

# Empirical examples

```{r tapir-example, fig.cap="Seven-day home range size of a tapir from Medici $et~al.$ (2022) as mean and variance in resource abundance change over the tracking period. (a.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir during the period of reference. (b.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir during the period of reference. (c.) Estimated home range size of the tapir during each reference period, based on 50% (bold) and 95% (thin) utilization distributions.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/tapir-example.png')
```

Environmental productivity is tightly linked to the amount of space animals need to cover to obtain the resources they needed to survive and reproduce [@lucherini_habitat_1996; @relyea_home_2000]. While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction, such as energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @refs], and maintaining a thermoregulatory balance. The size of a home range, is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [@nilsen_can_2005], since larger home ranges can result in higher rates of competition and are harder to defend [@grant_whether_1993; @jetz_scaling_2004].

There are many factors which may affect animal's spatial requirements and how they use their HR [*sensu* @nathan_movement_2008]. In particular, resource abundance and environmental productivity are often hypothesized to be inversely proportional to HR size (or some function of it), such that regions with higher abundance correspond to smaller spatial needs, since animals do not have to range over extensive areas to meet the energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @southwood_habitat_1977], and maintaining a thermoregulatory balance. The size of a home range is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear, since larger home ranges can result in higher rates of competition and are harder to defend [@grant_whether_1993; @jetz_scaling_2004]. @nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region. However, the effect of many other factors on HR size is likely depends on the species, location, and possibly also on the individual animal. For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; capuchin monkeys, @torrez-herrera_monkeys_2020]. Similarly, predation may force animals to move more frequently to escape predators, or it may prevent them from venturing too far from the safety of their core HR too often [@suraci_beyond_2022]. Patch quality, size, fragmentation, and heterogeneity may cause animals to explore more patches if some are of low value, too small, too disconnected, or too variable [@fahrig_is_2019], but high diversity may also decrease HR size if animals require heterogeneous habitats [@fox_niche_1981; @lucherini_habitat_1996]. Similarly, patch connectivity and ease of movement may widen HRs by decreasing the energetic cost of movement and favoring exploration [@dickie_resource_2022], or they may shrink HRs by decreasing the energetic cost of movement while increasing encounter rates with resources [@martinez-garcia_how_2020; @bartumeus_influence_2008; @visser_plankton_2006]. However, not all animals take advantage of linear features or higher predictability in human-altered habitats. @noonan_roads_2021 found that giant anteaters (*Myrmecophaga tridactyla*) in Brazil did not use roads to reduce movement costs or increase movement speed. Instead, roads increased anteater mortality because the animals were attracted to the high-quality foraging found on roadside habitat. Similarly, @medici_movement_2022 found that the movement of tapirs (*Tapirus terrestris*) was unaffected by anthropogenic activity or habitat type.

>>>>

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats 
    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    keep_tex: true
    number_sections: false
    toc: false
---

<!--
TO DO:
- write conclusion paragraph
- write discussion (2-3 pages)
    - add headers for sections
- trim down and edit intro and other previous sections
- move empirical examples within each relevant section (i.e., abundance examples for abundance section)
- simple figure for demonstrating idea of nonlinear relationship between R and H, with animal moving in empty space?
    - explain the idea in the caption, straight line can't work because can't have negative space, and there's need for saturation
    - line increasing as R decreases, 3 options: death, more increase (nomadism), sub-optimal saturation
- make same figure for V(R), but focus on temporal variance in time:
    - use grey or low alpha for curves with high uncertainty
- focus on V(R) as temporal variance rather than spatial one. people have already focused on patchiness, but temporal variation is more risky
- drop figure 1? (Fictitious example of variation)
- move 5x5 figure to interactive effects section
- add median NDVI behind tracking data of tapir, buffalo, and ??? figure
-->

\newpage <!-- start main body on a new page -->

@@

# chuck options used to wrap figures:
## fig.align = "right":          align figures right
## out.width = "0.5\\textwidth": output width = 0.5 * the width of the text
## wrapfigure = list("R", 0.5)   wrap caption to right, caption width = 0.5 text width

#' place figures before the paragraph they are referred to so `wrapfigure` places them
#' beside the paragraph
```

<!--
@ciuti and @berger-tal: human presence in nature for other reasons than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979
-->

# Abstract

Animal's spatial needs are often thought to be tighly linked to the amount of resources ($R$) within their habitats. Although this hypothesis has widespread support, most studies only account for the mean $R$, while the variance in $R$ is rarely accounted for. Failing to account for such variation is likely to result in an underestimation on animals' spatial needs, which may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict. In this paper, we provide a review of animal movement studies that estimate animals' spatial needs as a function of $R$, and we demonstrate the importance of accounting for variance in $R$. Additionally, we provide a unifying framework and hypotheses for the effect of mean and variance in $R$ on animals' home range sizes and movement behaviors. In the first part of this paper, we use simulated movement data in a variety of example habitats to demonstrate the importance of accounting for the variance in $R$. Next, we use the Normalized Difference Vegetation Index to estimate mean and variance in $R$ within the habitats of a tapir from the Pantanal (Brazil) and an African buffalo from South Africa to demostrate how this framework can be applied to real-world tracking data. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to real-world tracking data.

# Introduction

The ability to move allows motile organisms to respond rapidly and continuously to heterogeneous and changing environments and conditions [@southwood_habitat_1977]. Animals may move for a variety of reasons, including searching for resources [@refs] or mates [@refs] while avoiding predation [@refs] or dangerous competition [@refs], as well as defending resources and territories [@grant_whether_1993; @refs]. @nathan_movement_2008 conceptualized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

@@

such that the animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters, doing so with sufficient frequency and accuracy would not be simple. Rather than measuring changes in the various resources and needs directly, we can quantify animals' spatial needs as a response to average resource abundance and the variability around it, instead. This paper will illustrate how animals' spatial needs depend on average resource abundance and the variance around it. We present this idea through a quantitative theoretical framework and empirical examples where resource abundance is estimated using the Normalized Vegetation Index [NDVI, see @pettorelli_normalized_2011].

@@

Figure 2 of @bista_effects_2022 for example of study that assumes linear relationships (w LM) for home range, when there can't be one.

```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of mean resource abundance ($\\mathbb E(R)$). At low values of $\\mathbb E(R)$, range-resident animals (bottom dashed line) may exhibit nomadic behavior, such that While $H$ decreases as $\\mathbb E(R)$ increases, it is not possible for $H$ to continuously decrease linearly because animals will require a minimum (nonnegative) amount of space to survive. At low values of $\\mathbb E(R)$, $H$", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

Environmental productivity is tightly linked to the amount of space animals need to cover to obtain the resources they needed to survive and reproduce [figure \@ref(fig:), @relyea_home_2000]. While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction, such as energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @refs], and maintaining a thermoregulatory balance. The size of a home range is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005]. Animals living in large groups will require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], and larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004].

The favorableness of a patch or habitat often depends on a variety of factors, including resource abundance, competitive pressure, and predation risk. Since many of these factors' affect resource abundance unpredictably, it is useful to visualize it as a random variable^[In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$, $H$). Known values, such as realizations (i.e., known observations) of random variables, are indicated with lower-case letters (e.g., $r$, $h$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.]. Let $R$ indicate the amount of *resources* in a particular patch or area. For simplicity, we can assume $R$ can take any value between zero and infinity, and we can let $R$ follow a Gamma distribution with mean $\mu$ and variance $\sigma^2$ (which we can write as $R \sim \Gamma(\mu, \sigma^2)$)^[Although Gamma distributions are more often parameterized using shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ for simplicity. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean since as the mean approaches zero the variance also does: $\mu = 0 \iff k \theta = 0 \iff k \theta^2 = 0 \iff \sigma^2 = 0$.]. Next, let $H$ indicate the size of an animal's home range (HR)^[We can consider specific utilization quantiles of the HR, such as the core HR, $H_{50\%}$, or the 95% HR, $H_{95\%}$, but for simplicity I will refer to the entirety of the HR with $H$.].

@@

@@

The top row of figure \@ref(fig:5-5-sims) (constant $\mathbb V(R)$) illustrates the simulated spatial requirements of an animal living in habitats where $\mathbb E(R)$ remains constant or changes over time but $\mathbb V(R)$ remains constant. (Although some noise in $H$ are still present since $\mathbb V(R) \ne 0$ and thus the $R$ is not constant). Note how the home range size, $H$, decreases nonlinearly as $\mathbb E(R)$ increases linearly (linear $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), with larger changes when $\mathbb E(R)$ is low, since the multiplicative changes in $R$ are larger. Additionally, the 95% home range is more sensitive to changes in $\mathbb E(R)$ than the core home range $H_{50\%}$. In regions where the average resource abundance changes over time (cyclical $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mathbb E(R)$. However, estimates of spatial requirements based on estimated changes in $\mathbb E(R)$ should be interpreted carefully, since model error and unpredictable changes in $\mathbb E(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the expected spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), since accurate long-term estimates of $\mathbb E(R)$ may be hard to produce, if not impossible. In cases where $\mathbb E(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often. Finally, in ecosystems where $\mathbb E(R)$ changes erratically and suddenly (erratic $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time. The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile. Note how both quantiles decrease nonlinearly as $\\mathbb E(R)$ increases, and they increase approximately linearly as $\\mathbb V(R)$ increases. Additionally, the variance in both quantiles is higher when $\\mathbb V(R)$ is higher, and changes in $\\mathbb V(R)$ have greater impacts when $\\mathbb E(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time. The animal's spatial variance parameter was then calculated using an OUF model via the $\\tt{ctmm}$ package", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

@@

Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], the variance in $R$ has received far less attention. However, $R$ can vary strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018]. Environmental stochasticity, including extreme events, can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

@seigle-ferrand_systematic_2021: Landscape heterogeneity influenced the sizes of home ranges in about 96% of studies based on individuals monitored by VHF but in only about 75% for individuals monitored by GPS, whatever the number of studied metrics.

When animals are not guaranteed that the resources they find during one visit will be there the next time (figure \@ref(fig:multipanel-changing-var)a), $\mathbb V(R)$ will have an appreciable effect on the location's favourableness. Since $\mathbb V(R)$ is not constant over time, we can now define it as a function of time using the notation $\mathbb V(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b). In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).

In this paper, we separate $\mathbb V(R)$ into heterogeneity and stochasticity in $R$. We refer to predictable or expected spatiotemporal changes as environmental **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable. In contrast, whether it is a good or bad year for berries is stochastic. <!--Thus, changes in $\mathbb V(R)$ may predictable or not, and $\mathbb V(R)$ is affected by both sources of variation.-->

The leftmost column of figure \@ref(fig:5-5-sims) (constant $\mathbb E(H)$) illustrates how an animal's home range may vary as $\mathbb V(R)$ changes over time while $\mathbb E(R)$ remains constant. Both $\mathbb E(H)$ and $\mathbb V(H)$ increase with $\mathbb V(R)$, although $H_{95\%}$ is more sensitive to changes in $\mathbb V(R)$ than $H_{50\%}$. Additionally, the effect of $\mathbb V(R)$ on $H$ depends on $\mathbb E(R)$, since the effect of $\mathbb V(R)$ on $H$ is stronger when $\mathbb E(R)$ is low (see the bottom row with erratic $\mathbb V(R)$).

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\mathbb E(R)$ and $\mathbb V(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\mathbb E(R)$ and $\mathbb V(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\mathbb E(R)$ and $\mathbb V(R)$ remain undetected. Although it is impossible for $\mathbb E(R)$ and $\mathbb V(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease). Additionally, these examples are important because they demonstrate the relationships between $H$, $\mathbb E(R)$, and $\mathbb V(R)$ in (relatively) simple scenarios. Cyclical oscillations in $\mathbb E(R)$ and $\mathbb V(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\mathbb E(R)$ and $\mathbb V(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity. Finally, erratic changes in $\mathbb E(R)$ and $\mathbb V(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\mathbb E(R)$ or increases in $\mathbb V(R)$ rather a series of small, sudden, changes. Estimating the true trend in $\mathbb E(R)$ and $\mathbb V(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\mathbb E(R)$ are not due to $\mathbb V(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\mathbb E(R)$ and $\mathbb V(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$. <!-- https://doi.org/10.1080/00207549508930162 ?-->

The effects of $\mathbb V(R)$ on animals' spatial requirements likely depend on the animals' needs. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic. In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions. Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality. The work by @nilsen_can_2005 also demonstrates that the effect of $\mathbb E(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\mathbb V(R)$, so it is important to account for interaction effects between the two.

What an animal perceives as a stochastic process depends on the stochasticity of the process relative to the animal's size, current age (or average lifespan), and adaptability. Generally, small, short-lived, or young organisms will tend to be more sensitive to small-scale (spatial) or short-term (temporal) changes [@southwood_habitat_1977]. Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004], which allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high fertility of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can allow them to develop traits that increase the chances of survival following an extreme event. Additionally, the effects of size and lifespan on sensitivity are likely nonlinear and correlated, since smaller animals tend to have shorter lives (and vice-versa), and small animals also tend to have lower metabolic rates, which often limit their movement speed, home ranges, and how the animals interact with their ecosystem [@brown_toward_2004]. Thus, processes will have stronger impacts on smaller animals than on bigger ones: The grazing pattern of a bison (*Bison bison*) drastically alters the habitat of most crawling insects, but other grazing mammals would not pay attention to changes in grass length at the same spatial or temporal resolution. Similarly, the timing and quantity of yearly snowfall would be a somewhat predictable and expected event for most adult moose (*Alces alces*), but it may be a shock for many adult zooplankton or a new-born wolf (*Canis lupus*). What one animal may perceive as a single stochastic event (or cycle) may be considered as a series of short and highly stochastic events by another animal. Therefore, the mean and variance in $R$ should be modeled while accounting for the behavior of the species of interest.

An unresponsive animal may perceive frequent changes in $R$ as high $\mathbb V(R)$, while a responsive animal may perceive them as changes in $\mathbb E(R)$. Thus, $R$ should be modeled using very flexible and wiggly models when animals are highly responsive, while less flexible models are more appropriate for more unresponsive or sedentary species. Generalized Additive Models [GAMs, see @wood_generalized_2017] are sufficiently flexible and adaptable to allow one to model $R$ while reflecting various levels of adaptability and responsiveness. (See @simpson_modelling_2018 for an overview of how to select appropriate bases sizes and smoothness parameters for GAMs.) The animal's movement (e.g., degree of range residency, range crossing time, average speed, directional persistence) should also inform the complexity of any spatial smooths included in the models. Models with complex spatial terms are appropriate for animals that recognize the location of high-quality patches, while smoother terms are best for animals that may perceive differences in $R$ as spatial variance, instead.

**it may be worth chatting about the NDVI models for the tapir and the buffalo later this month**

# Discussion
